---
title: "Basic statistics for health research - week 1, ex. 1"
author: "Benjamin Skov Kaas-Hansen"
date: "9/9/2021"
output: 
  html_document: 
    code_folding: show
    df_print: tibble
    highlight: haddock
    theme: paper
    toc: yes
    toc_depth: 3
---

### 0. Setup

Load packages
```{r setup, warning=FALSE, message=FALSE}
# Set default alignment of figures
knitr::opts_chunk$set(fig.align = "center")

# Data manipulation
library(dplyr)
library(tidyr)
library(readr)

# Data inspection
library(mice)

# Data visualiation
library(ggplot2)
library(qqplotr) # for QQ-plots, a ggplot2 extension
library(patchwork) # for patching together plots, very powerful and simple to use (highly recommended)
theme_set(theme_minimal()) # sets default theme for ggplots
```

Define helper functions (if you do the same or almost the same thing more than twice, you should create a function for that specific task.) `tribble` is a way to create dataframes row-wise; `pivot_longer` and `pivot_wider` come from the `tidyr` package and allows for converting between long and wide format (more on this later in the course).

```{r}
summary_stats <- function(x) {
	"
	x: vector of continuous values
	"
	tribble(
		~ statistic, ~ value,
		"mean", mean(x, na.rm = TRUE),
		"std_dev", sd(x, na.rm = TRUE),
		"median", median(x, na.rm = TRUE),
		"p25", quantile(x, probs = 0.25, na.rm = TRUE),
		"p75", quantile(x, probs = 0.75, na.rm = TRUE),
		"min", min(x, na.rm = TRUE),
		"max", max(x, na.rm = TRUE),
		"n_missing", sum(is.na(x))
	)
}

pivot_around_sex <- function(df) {
	"
	df is a data frame with at least a column named sex and some data columns to pivot
	"
	df %>% 
		pivot_longer(cols = -sex, names_to = "statistic", values_to = "value") %>% 
		pivot_wider(names_from = "sex")
}
```

### 1. Data wrangling
Remember that `transmute` discards all columns not modified or referenced ("mentioned") in the call. It's an easy to do wrangling + renaming of incoming data while throwing away the old columns. I prefer to keep all data wrangling together, but you find out as you go what you need: it's an iterative process, and so it makes sense to come back and add the data wrangling here when you realise you need something (e.g. BMI in this example).

```{r data_wrangling}
df <- read_delim("http://staff.pubhealth.ku.dk/~sr/BasicStatistics/datasets/sundby.txt", delim = " ", na = ".") %>% 
	transmute(weight = v75,
			  height = v76 / 100, # cm ~> metres
			  sex = factor(kon, levels = c(1, 2), labels = c("male", "female")),
			  bmi = weight / height^2)

glimpse(df) # gives a nice overview of the actual values in the dataframe ≠ summary()
```

Get a few summary stats; we have `r nrow(df)` observations with `r ncol(df)` variables.
```{r}
summary(df)
```

#### Pattern in missingness
The rows are missingness pattern (0 = variable is missing). The left-most column is the number of patients with the respective missingness patters, the right-most column the number of variables missing in each pattern, the bottom-most row the number of participants with missing values for the respective variables.

```{r}
md.pattern(df, plot = FALSE)
```

### 2. Weights of men and women
The weight distributions seem to differ although they do overlap quite a lot.

#### Numerical summary
```{r}
df %>% 
	group_by(sex) %>% 
	summarise(summary_stats(weight)) %>% # note that summarise() here actually reduces to >1 row
	arrange(statistic, sex)
```

#### Plotting

QQ-plot
```{r}
ggplot(df, aes(sample = weight, colour = sex)) +
	geom_abline(intercept = 0, slope = 1, size = 0.25) + # the name comes from the equation: y = ax + b
	stat_qq_point(size = 0.5) # from qqplotr
```


First a histogram of the weight distribution overall. Note: different values of `binwidth` will change the appearance quite a lot (as all histograms)
```{r}
ggplot(df, aes(x = weight)) +
	geom_histogram(binwidth = 5)
```

We often want to show men and women separately. The "naive" approach seems weird because `ggplot2` stacks histograms by default.
```{r}
ggplot(df, aes(x = weight)) +
	geom_histogram(aes(fill = sex), binwidth = 5)
```

So we tell `ggplot2` to dodge (= "unstack") the bars
```{r}
ggplot(df, aes(x = weight)) +
	geom_histogram(aes(fill = sex), binwidth = 5, position = "dodge")
```

It's definitely better, but it can be a bit tricky to compare distributions without the same number of observations because the modes won't be of equal height. Using `table` we can see that there are about 21% more women than men.
```{r}
table(df$sex) # many more females, so a crude histogram might be misleading
```

--so we want "normalised" histograms instead. `ggplot2` computed some values (which ones depends on the `geom`), and these can be called surrounding them by two periods (see e.g. `?geom_histogram`)
```{r}
ggplot(df, aes(x = weight)) +
	geom_histogram(aes(fill = sex, y = ..ncount..), binwidth = 5, position = "dodge")
```

I think multiple histrograms together are ugly and prefer to plot the frequencies as lines instead (= frequency polygons in `ggplot2`; weird name, pretty plot). I add some plot cosmetics to have a nice-looking and near publication-ready plot:
```{r}
ggplot(df, aes(x = weight, colour = sex, y = ..ncount..)) +
	geom_freqpoly(binwidth = 5, size = 0.25) +
	stat_bin(size = 0.5, binwidth = 5, geom = "point", position = "identity") +
	labs(x = "Weight in kilograms", y = "No. participants as fraction of maximum count") +
	theme(axis.text.y = element_blank())
```

#### Alternatives
Violin plots are richer than boxplots and can unearth patterns in the distribution that escape boxplots. Their appearance depends heavily on their settings so be careful. 

```{r}
ggplot(df, aes(x = sex, y = weight, fill = sex)) +
	geom_violin(size = 0.4) 

ggplot(df, aes(x = sex, y = weight, fill = sex)) +
	geom_boxplot(size = 0.4) 
```

### 3. Construct reference intervals

#### On the "circa 2" factor
Often, you can just use 2 or 1.96 (because the $t$ distribution gets closer to the Normal as N becomes large). This is just to show the relationship between the "circa 2"-factor and the number of observations. `tibble`s are a modern version of dataframes (they arrived with `dplyr`) and offer three nice features: columns are created "on-the-fly" so you can use values from previous columns to create a new one (this is not possible with `data.frame()`), they print in a nice way (unlike dataframes; this becomes more important for larger dataframes/tibbles), they don't convert strings to factors automatically (the `read_*()` functions from `readr` have the same feature unlike the `read.*()` in base R). As we can see, already for around 100 observations, the 1.96 approximation is quite good.
```{r}
tibble(n_obs = 2:200, # n_obs = 1 is meaningless as a single observation can't have a distribution
	   circa_2 = (1 + 1/n_obs) * qt(0.975, n_obs - 1)) %>% 
	ggplot(aes(x = n_obs, y = circa_2)) +
		geom_hline(yintercept = qt(0.975, Inf), colour = "orange", linetype = 2, size = 0.5) +
		geom_line(size = 0.2) +
		geom_point(size = 0.2) +
		scale_x_log10() +
		scale_y_continuous(breaks = c(round(qt(0.975, Inf), 2), 5, 10, 15, 20))
```

#### Using the Normality assumption
```{r}
df %>% 
	filter(!is.na(weight)) %>% # keep the rows where weight is *not* missing
	group_by(sex) %>%
	summarise(mean = mean(weight),
			  std_dev = sd(weight),
			  interval_factor = sqrt(1 + 1/n()) * qt(0.975, n() - 1),
			  ci_lo_norm = mean - interval_factor * std_dev,
			  ci_hi_norm = mean + interval_factor * std_dev,
			  ci_lo_nonnorm = quantile(weight, 0.025),
			  ci_hi_nonnorm = quantile(weight, 0.975)) %>% 
	pivot_around_sex() # this is just cosmetics so the results fit better
```

### 5. Scatterplot of weight ~ height
The simple scatterplot looks nice but is actually deceiving.
```{r}
ggplot(df, aes(x = weight, y = height, shape = sex, colour = sex)) +
	geom_point()
```

If we make the points transparent, we see many points are plotted on top of each other (over-plotting) which means the first opaque scatterplot is "hiding" observations. Making points semi-transparent helps (`alpha = 1` is fully opaque, `alph = 0` is fully transparent):
```{r}
ggplot(df, aes(x = weight, y = height, shape = sex, colour = sex)) +
	geom_point(alpha = 0.5)
```

But we can also try a heatmap-plot (note again the use of a computed value surrouned by two periods, here `..ncount..`)
```{r}
ggplot(df, aes(x = weight, y = height)) +
	geom_hex(aes(alpha = ..ncount.., fill = sex), bins = 40)
```

A contour plot is perhaps less confusing than the heatmap plot because overlaying of colours isn't handled well.
```{r}
ggplot(df, aes(x = weight, y = height, colour = sex)) +
	geom_density_2d(size = 0.25) 
```

### 6. BMI

The distribution seems about right: right-skewed, with some extreme values but mostly in the normal range. BMI doesn't seem to have a Normal distribution, however. I'm overlaying the plot with the density of the Normal distribution that we assume based on the mean and standard deviation (blue curve). If this is too difficult, just dont' try to understand the `stat_function` line and we'll come back to it later in the course. 

```{r}
df_bmi <- df %>% 
	filter(!is.na(bmi))

ggplot(df_bmi, aes(x = bmi)) +
	geom_histogram(aes(y = ..density..), colour = "white", fill = "grey80", binwidth = 1) +
	geom_density() +
	stat_function(colour = "dodgerblue", fun = dnorm, args = list(mean = mean(df_bmi$bmi), sd = sd(df_bmi$bmi))) 
```

The QQ-plot is also off
```{r}
ggplot(df_bmi, aes(sample = bmi)) +
	geom_abline(intercept = 0, slope = 1) +
	stat_qq_point()
```

However, it might be misleading to plot men and women together:
```{r}
df_men <- df_bmi %>% 
	filter(sex == "male") 
df_women <- df_bmi %>% 
	filter(sex == "female")

p_men <- ggplot(df_men, aes(x = bmi)) +
	geom_histogram(aes(y = ..density..), colour = "white", fill = "grey80", binwidth = 1) +
	geom_density() +
	stat_function(colour = "dodgerblue", fun = dnorm, args = list(mean = mean(df_men$bmi), sd = sd(df_men$bmi))) +
	labs(title = "Men")

p_women <- ggplot(df_women, aes(x = bmi)) +
	geom_histogram(aes(y = ..density..), colour = "white", fill = "grey80", binwidth = 1) +
	geom_density() +
	stat_function(colour = "dodgerblue", fun = dnorm, args = list(mean = mean(df_women$bmi), sd = sd(df_women$bmi))) +
	labs(title = "Women")

p_men + p_women # the patchwork package allows us to patch together plots like this
```

Or as the frequency polygon instead:
```{r}
ggplot(df_bmi, aes(x = bmi, colour = sex)) +
	geom_freqpoly(binwidth = 1)
```

#### Determine normal ranges for each sex with and without normality assumption
Note that use of `sum` and `mean` of TRUE/FALSE vectors (whether BMI is between the lower and higer bounds of the reference intervals.)

```{r}
df_bmi %>%
	group_by(sex) %>%
	summarise(mean = mean(bmi),
			  std_dev = sd(bmi),
			  interval_factor = sqrt(1 + 1/n()) * qt(0.975, n() - 1), # = circa 2
			  ci_lo_norm = mean - interval_factor * std_dev,
			  ci_hi_norm = mean + interval_factor * std_dev,
			  n_outside_ci_norm = sum(!between(bmi, ci_lo_norm, ci_hi_norm)),
			  prop_outside_ci_norm = mean(!between(bmi, ci_lo_norm, ci_hi_norm)),
			  ci_lo_nonnorm = quantile(bmi, 0.025),
			  ci_hi_nonnorm = quantile(bmi, 0.975),
			  n_outside_ci_nonnorm = sum(!between(bmi, ci_lo_nonnorm, ci_hi_nonnorm)),
			  prop_outside_ci_nonnorm = mean(!between(bmi, ci_lo_nonnorm, ci_hi_nonnorm))) %>% 
	pivot_around_sex() # again, just cosmetics
```

### 7. Using log-BMI
Better agreement between intervals based on normality assumption and empirical estimates. 
```{r}
df_bmi %>% 
	mutate(log_bmi = log(bmi)) %>% 
	group_by(sex) %>%
	summarise(mean = mean(log_bmi),
			  std_dev = sd(log_bmi),
			  interval_factor = sqrt(1 + 1/n()) * qt(0.975, n() - 1),
			  ci_lo_norm = exp(mean - interval_factor * std_dev),
			  ci_hi_norm = exp(mean + interval_factor * std_dev),
			  n_outside_ci_norm = sum(!between(bmi, ci_lo_norm, ci_hi_norm)),
			  prop_outside_ci_norm = mean(!between(bmi, ci_lo_norm, ci_hi_norm)),
			  ci_lo_nonnorm = exp(quantile(log_bmi, 0.025)),
			  ci_hi_nonnorm = exp(quantile(log_bmi, 0.975)),
			  n_outside_ci_nonnorm = sum(!between(bmi, ci_lo_nonnorm, ci_hi_nonnorm)),
			  prop_outside_ci_nonnorm = mean(!between(bmi, ci_lo_nonnorm, ci_hi_nonnorm))) %>% 
	pivot_around_sex()
```
