---
title: "Basic statistics for health research - week 7, ex. 1"
author: "Benjamin Skov Kaas-Hansen"
date: "28/10/2021"
output: 
  html_document: 
    theme: paper
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

### Prep and setup
```{r}
library(plyr) # must be loaded before dplyr (which is loaded with tidyverse)
library(tidyverse)
library(purrr)
library(lindia)

theme_set(theme_minimal())

#' @param m model object
#' returns a matrix
pretty_estimates <- function(m) { 
	cbind(Est. = m$coefficients, confint(m))
}
```

# Cadmium (+ Q1)
```{r}
cad <- read_delim("http://publicifsv.sund.ku.dk/~lts/basal/data/cadmium.txt", delim = " ") %>% 
	mutate(grp = factor(grp, levels = 1:3, labels = c("long_exposure", "short_exposure", "no_exposure")))
```

## Q2: Group-specific summaries
High-exposure individuals seems different (lower vital capacity, higher age), but not substantively so. Vital capacity decreases with age, so that's a quite likely explanation. 

```{r}
dlply(cad, "grp", summary)

# Alternative
split(cad, cad$grp) %>% 
	map(summary)

cad %>% 
	pivot_longer(-grp, names_to = "var", values_to = "value") %>% 
	ggplot(aes(x = grp, y = value, colour = grp)) +
		geom_boxplot(position = "dodge2", fill = NA, show.legend = FALSE) +
		facet_wrap(~ var, scales = "free") +
		theme(panel.spacing = grid::unit(2, "lines"),
			  panel.grid.major.x = element_blank()) +
		labs(x = "", y = "Variable value")
```

## Q3: Difference between vital capacity across the exposure groups?
```{r}
model1 <- lm(vitcap ~ grp, data = cad)
summary(model1)
round(pretty_estimates(model1), 2)
```

The F test has a p-value of 0.09 so not significant difference, but would you really rule it out entirely? Also because the p values for the coefficients from the two exposure groups are actually below 0.05. It's important, however, to note that the CI's for these confidence intervals are basically 0 to 1, so quite imprecise. 

If we want to study the pairwise differences, we have to use Tukey-corrected T tests. That requires an analysis-of-variance, and running the Tukey multiple comparisons of means analysis we see that all CI's contain 0 (although not by a large margin for short vs. long exposure and no vs. long exposure, which corresponds well to what we saw in the boxplots). Also note that the CI's are now wider because we've corrected for multiple comparisons.

```{r}
model2 <- aov(vitcap ~ grp, data = cad)
TukeyHSD(model2)
```

### Diagnostic plots
```{r, fig.height=10}
gg_diagnose(model1)

with(cad, bartlett.test(vitcap, grp)) # test of variance homogeneity
kruskal.test(vitcap ~ grp, data = cad) # in case of violation of the assumption of normally-distributed residuals
```

## Q4: Correlation between vitcap and age within each group
I'll do linear model instead of correlation test per se. The regression coefficients within each group might sugggest that vital capacity fall more rapidly in the long-term exposed individuals. Inteterstingly, the long-term exposed have higher vitcap when younger. Note also that ggplot2 doesn't extrapolate the lines beyond the data, as base graphics in R do.
```{r}
fits <- list(overall = lm(vitcap ~ age, data = cad)) %>% 
	c(dlply(cad, "grp", function(d) lm(vitcap ~ age, data = d)))

llply(fits, summary) # requires plyr
map(fits, summary) # alternative with purrr

llply(fits, pretty_estimates) # requires plyr
map(fits, pretty_estimates) # alternative with purrr

ggplot(cad, aes(age, vitcap, colour = grp)) +
	geom_point() +
	geom_smooth(method = "lm", formula = y ~ x, se = FALSE, size = 0.4) +
	geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = FALSE, linetype = 2, size = 0.4)
```

The plot clearly shows that the vital capacity decreases more rapidly in long-term exposed individuals compared to the others. Also, the not exposed have the slowest deterioration. The GAM lines above are not very different from the straight lines.

## Q6: Provide estimate for the difference between 40-year old long-term exposed individuals and 40-year old non-exposed
We need a combined model (of course). From the ANOVA we see that there are differences in vitcap for 40-year old individuals across the groups (p = 0.034), but we don't know how. The difference between the two groups can not be read directly from the coefficients (between we centered age at 40 years), and suggests that long-term exposure is favourable to non-exposure (negative coefficient for being non-exposed); it isn't significant, though. When looking at the plot above, we can also see that making statements on 40-year old long-term exposed individuals is a bit sketchy as we are at the bound of the data. 
```{r}
cad <- cad %>% 
	mutate(age_around_40 = age - 40)
model3 <- lm(vitcap ~ age_around_40 * grp, data = cad) 
# model3 <- lm(vitcap ~ I(age - 40) * grp, data = cad) # alternative, but gives ugly coefficient names
summary(model3)
pretty_estimates(model3)
anova(model3) # shows significant interaction, so it should stay in the model

# Difference between estimates from two models
bind_cols(model = c("Only group as covariate", "Interaction b/w age and grp"),
		  bind_rows(pretty_estimates(model3)[4, ], pretty_estimates(model1)[3, ]))
```

## Q7: How many more years does it take to lose 1 litre of vital capacity when one is non-exposed compared with long-term exposed?
We just want a point estimate, so use the slope (unit: liters lost per year) and use the reciprocal to get years to one liter. We are looking for the difference between these two values (approx. 20.9 years)
```{r}
1/model3$coefficients["age_around_40"]
1/(model3$coefficients["age_around_40"] + model3$coefficients["age_around_40:grpno_exposure"]) # = 1/0.03061
```
