---
title: "Uge 6 notebook"
author: "Benjamin Skov Kaas-Hansen"
date: "10/7/2020"
output: 
  html_document: 
    theme: paper
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

### Prep and setup
```{r}
for (p in (c("plyr", "tidyverse")))
	library(p, character.only = TRUE)
theme_set(theme_minimal())

pretty_estimates <- function(m) { # m: model object
	cbind(Est. = m$coefficients, confint(m))
}
```

# Cadmium (+ Q1)
```{r}
cad <- read.table("http://publicifsv.sund.ku.dk/~lts/basal/data/cadmium.txt", header = TRUE) %>% 
	mutate(grp = factor(grp, levels = 1:3, labels = c("long_exposure", "short_exposure", "no_exposure")))
```

## Q2: Group-specific summaries
High-exposure individuals seems different (lower vital capacity, higher age), but not substantively so. Vital capacity decreases with age, so that's a quite likely explanation. 

```{r}
dlply(cad, "grp", summary)

gather(cad, var, value, -grp) %>% 
	ggplot(aes(x = grp, y = value, colour = grp)) +
		geom_boxplot(position = "dodge2", fill = NA, show.legend = FALSE) +
		facet_wrap(~ var, scales = "free") +
		theme(panel.spacing = grid::unit(2, "lines"),
			  panel.grid.major.x = element_blank()) +
		labs(x = "", y = "Variable value")
```

## Q3: Difference between vital capacity across the exposure groups?
```{r}
model1 <- lm(vitcap ~ grp, data = cad)
summary(model1)
round(pretty_estimates(model1), 2)
```

The F test has a p-value of 0.09 so not significant difference, but would you really rule it out entirely? Also because the p values for the coefficients from the two exposure groups are actually below 0.05. It's important, however, to note that the CI's for these confidence intervals are basically 0 to 1, so quite imprecise. 

If we want to study the pairwise differences, we have to use Tukey-corrected T tests. That requires an analysis-of-variance, and running the Tukey multiple comparisons of means analysis we see that all CI's contain 0 (although not by a large margin for short vs. long exposure and no vs. long exposure, which corresponds well to what we saw in the boxplots). Also note that the CI's are now wider because we've corrected for multiple comparisons.

```{r}
model2 <- aov(vitcap ~ grp, data = cad)
TukeyHSD(model2)
```

### Diagnostic plots
```{r}
plot(model1, which = c(1, 2, 4))
hist(resid(model1))
with(cad, bartlett.test(vitcap, grp)) # test of variance homogeneity
kruskal.test(vitcap ~ grp, data = cad) # in case of violation of the assumption of normally-distributed residuals
```

## Q4: Correlation between vitcap and age within each group
I'll do linear model instead of correlation test per se. The regression coefficients within each group might sugggest that vital capacity fall more rapidly in the long-term exposed individuals. Inteterstingly, the long-term exposed have higher vitcap when younger. Not also that ggplot2 doesn't extrapolate the lines beyond the data, as base graphics in R do.
```{r}
fits <- list(overall = lm(vitcap ~ age, data = cad)) %>% 
	c(dlply(cad, "grp", function(d) lm(vitcap ~ age, data = d)))
llply(fits, summary)
llply(fits, pretty_estimates)

ggplot(cad, aes(age, vitcap, colour = grp)) +
	geom_point() +
	geom_smooth(method = "lm", formula = y ~ x, se = FALSE, size = 0.4) +
	geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = FALSE, linetype = 2, size = 0.4)
```

The plot clearly shows that the vital capacity decreases more rapidly in long-term exposed individuals compared to the others. Also, the not exposed have the slowest deterioration. The GAM lines above are not very different from the straight lines.


## Q6: Provide estimate for the difference between 40-year old long-term exposed individuals and 40-year old non-exposed
We need a combined model (of course). From the ANOVA we see that there are differences in vitcap for 40-year old individuals across the groups (p = 0.034), but we don't know how. The difference between the two groups can not be read directly from the coefficients (between we centered age at 40 years), and suggests that long-term exposure is favourable to non-exposure (negative coefficient for being non-exposed); it isn't significant, though. When looking at the plot above, we can also see that making statements on 40-year old long-term exposued individuals is a bit sketchy as we are at the bound of the data. 
```{r}
cad <- mutate(cad, age_around_40 = age - 40)
model3 <- lm(vitcap ~ age_around_40 * grp, data = cad) 
summary(model3)
pretty_estimates(model3)
anova(model3) # shows significant interaction, so it should stay in the model

# Difference between estimates from two models
bind_cols(model = c("Only group as covariate", "Interaction b/w age and grp"),
		  bind_rows(pretty_estimates(model3)[4, ], pretty_estimates(model1)[3, ]))
```

# Growth hormone exercise
## Q1
```{r}
juul <- read.table("http://publicifsv.sund.ku.dk/~lts/basal/data/juul2.txt", header = TRUE, na = ".") %>% 
	mutate(sex = factor(sexnr, levels = 1:2, labels = c("Male", "Female")),
		   log_sigf1 = log(sigf1))

model_females <- lm(log_sigf1 ~ age, data = juul, subset = sex == "Female" & tanner == 1)
summary(model_females)
exp(pretty_estimates(model_females))

model_males <- update(model_females, subset = sex == "Male" & tanner == 1)
summary(model_males)
exp(pretty_estimates(model_males)) # exponentiate to get effect estimate (fraction change)

ggplot(filter(juul, tanner == 1), aes(age, log_sigf1, colour = sex)) +
	geom_point(alpha = 0.5) +
	geom_smooth(method = "lm", formula = y ~ x, se = FALSE, size = 0.5)

# Percent change in 5 years (disregard intercept here, as it doesn't change over time)
exp(5 * pretty_estimates(model_females))
exp(5 * pretty_estimates(model_males))
```

## Q2: Are the regression lines parallel?
Make model with interaction. The interaktion near-significant, so we cannot really rule out non-parallel lines. 
```{r}
model_int <- lm(log_sigf1 ~ age * sex, data = juul, subset = tanner == 1)
summary(model_int)
pretty_estimates(model_int)

juul <- mutate(juul, age_around_7 = age - 7) # same trick as above to get differences directly
model_int7 <- lm(log_sigf1 ~ age_around_7 * sex, data = juul, subset = tanner == 1)
summary(model_int7)
pretty_estimates(model_int7)
```

It seems that girls, at the age of 7, have about 21% (9-35) higher levels of IGF1: 
```{r}
exp(pretty_estimates(model_int7))
```

### If we assume age effect is the same for both sexes, we leave out of the interaction term
This shows a strong effect of both age and sex, which isn't really surprising.
```{r}
ancova <- lm(log_sigf1 ~ sex + age, data = juul, subset = tanner == 1)
summary(ancova)
pretty_estimates(ancova)
```

To visualise these parallel lines, we now have to make predictions and show those as a line:
```{r}
preds <- expand.grid(sex = c("Female", "Male"), age = 0:13, stringsAsFactors = FALSE) %>% 
	bind_cols(pred = predict(ancova, .))

ggplot(mapping = aes(x = age, colour = sex)) +
	geom_point(aes(y = log_sigf1), filter(juul, tanner == 1)) +
	geom_line(aes(y = pred), preds)
```

## Q3: Why would linear regression be inappropraite for all Tanner groups?
Just show the scatter plots and a smooth line; it's in no way straight.
```{r}
select(juul, log_sigf1, age, sex) %>% 
	na.exclude() %>% 
	ggplot(aes(age, log_sigf1)) +
		geom_point() +
		geom_smooth() +
		facet_wrap(~ sex)
```

## Q4: Consider only post-pubertal
For males, linear effect seems plausible but definitely not for women.
```{r}
select(juul, log_sigf1, age, sex) %>% 
	na.exclude() %>% 
	filter(age > 25) %>% 
	ggplot(aes(age, log_sigf1)) +
		geom_point() +
		geom_smooth(se = FALSE, colour = "red") +
		facet_wrap(~ sex)

select(juul, log_sigf1, age, sex) %>% 
	na.exclude() %>% 
	filter(age > 25) %>% 
	ggplot(aes(age, log_sigf1, colour = sex)) +
		geom_point() +
		geom_smooth(method = "lm", se = FALSE)
```

Fit model with interaktion as the lines do not seem parallel. The interaction term however is far from significant, and the effect is near-zero.
```{r}
model_postpub_int <- lm(log_sigf1 ~ sex*age, data = juul, subset = age > 25)
summary(model_postpub_int)
pretty_estimates(model_postpub_int)
```

--so we remove the interaction term and find the age effect to be significant unlike the sex effect. Not the effect direction: IGF1 increases with age in the young, and decreases with age in adult.

```{r}
ancova_postpub <- update(model_postpub_int, formula = . ~ sex + age)
summary(ancova_postpub)
pretty_estimates(ancova_postpub)
```

Again to visualise these parallel lines, we need to predict with the model:
```{r}
preds <- expand.grid(sex = c("Female", "Male"), age = 25:82, stringsAsFactors = FALSE) %>% 
	bind_cols(pred = predict(ancova_postpub, .))

ggplot(mapping = aes(x = age, colour = sex)) +
	geom_point(aes(y = log_sigf1), filter(juul, age > 25)) +
	geom_line(aes(y = pred), preds)
```

## Q5: Include BMI
```{r}
juul <- mutate(juul, bmi = weight / (height / 100)^2)
glm <- lm(log_sigf1 ~ age + sex + bmi, data = juul, subset = age > 25)
summary(glm) # note that 96 observation are deleted to missingness!
pretty_estimates(glm)
```

Always look at the missigness pattern! Note that I transpose it, as I find it to give a more intuitive output. 1’s mean observed value, 0’s mean missing value. Column titles indicate the number of patients with the pattern in that column. The bottom row shows the total number of variables with missing values in that column. The right-most column shows the number of patients with missing values for each variable. This is not the case for the plot which is in the original direction. 
```{r}
library(mice)
t(md.pattern(juul, rotate.names = TRUE))
```

