---
title: "Basic statistics for health research - week 7, ex. 2"
author: "Benjamin Skov Kaas-Hansen"
date: "25/3/2021"
output: 
  html_document: 
    theme: paper
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

### Prep and setup
```{r}
library(plyr)
library(tidyverse)

theme_set(theme_minimal())

pretty_estimates <- function(m) { # m: model object
	cbind(Est. = m$coefficients, confint(m))
}
```

# Growth hormone exercise
## Q1
```{r}
juul <- read.table("http://publicifsv.sund.ku.dk/~lts/basal/data/juul2.txt", header = TRUE, na = ".") %>% 
	mutate(sex = factor(sexnr, levels = 1:2, labels = c("Male", "Female")),
		   log_sigf1 = log(sigf1))

model_females <- lm(log_sigf1 ~ age, data = juul, subset = sex == "Female" & tanner == 1)
summary(model_females)
exp(pretty_estimates(model_females))

model_males <- update(model_females, subset = sex == "Male" & tanner == 1)
summary(model_males)
exp(pretty_estimates(model_males)) # exponentiate to get effect estimate (fraction change)

ggplot(filter(juul, tanner == 1), aes(age, log_sigf1, colour = sex)) +
	geom_point(alpha = 0.5) +
	geom_smooth(method = "lm", formula = y ~ x, se = FALSE, size = 0.5)

# Percent change in 5 years (disregard intercept here, as it doesn't change over time)
exp(5 * pretty_estimates(model_females))
exp(5 * pretty_estimates(model_males))
```

## Q2: Are the regression lines parallel?
Make model with interaction. The interaktion near-significant, so we cannot really rule out non-parallel lines. 
```{r}
model_int <- lm(log_sigf1 ~ age * sex, data = juul, subset = tanner == 1)
summary(model_int)
pretty_estimates(model_int)

juul <- mutate(juul, age_around_7 = age - 7) # same trick as above to get differences directly
model_int7 <- lm(log_sigf1 ~ age_around_7 * sex, data = juul, subset = tanner == 1)
summary(model_int7)
pretty_estimates(model_int7)
```

It seems that girls, at the age of 7, have about 21% (9-35) higher levels of IGF1: 
```{r}
exp(pretty_estimates(model_int7))
```

### If we assume age effect is the same for both sexes, we leave out of the interaction term
This shows a strong effect of both age and sex, which isn't really surprising.
```{r}
ancova <- lm(log_sigf1 ~ sex + age, data = juul, subset = tanner == 1)
summary(ancova)
pretty_estimates(ancova)
```

To visualise these parallel lines, we now have to make predictions and show those as a line:
```{r}
preds <- expand.grid(sex = c("Female", "Male"), age = 0:13, stringsAsFactors = FALSE) %>% 
	bind_cols(pred = predict(ancova, .))

ggplot(mapping = aes(x = age, colour = sex)) +
	geom_point(aes(y = log_sigf1), filter(juul, tanner == 1)) +
	geom_line(aes(y = pred), preds)
```

## Q3: Why would linear regression be inappropraite for all Tanner groups?
Just show the scatter plots and a smooth line; it's in no way straight.
```{r}
select(juul, log_sigf1, age, sex) %>% 
	na.exclude() %>% 
	ggplot(aes(age, log_sigf1)) +
		geom_point() +
		geom_smooth() +
		facet_wrap(~ sex)
```

## Q4: Consider only post-pubertal
For males, linear effect seems plausible but definitely not for women.
```{r}
select(juul, log_sigf1, age, sex) %>% 
	na.exclude() %>% 
	filter(age > 25) %>% 
	ggplot(aes(age, log_sigf1)) +
		geom_point() +
		geom_smooth(se = FALSE, colour = "red") +
		facet_wrap(~ sex)

select(juul, log_sigf1, age, sex) %>% 
	na.exclude() %>% 
	filter(age > 25) %>% 
	ggplot(aes(age, log_sigf1, colour = sex)) +
		geom_point() +
		geom_smooth(method = "lm", se = FALSE)
```

Fit model with interaktion as the lines do not seem parallel. The interaction term however is far from significant, and the effect is near-zero.
```{r}
model_postpub_int <- lm(log_sigf1 ~ sex*age, data = juul, subset = age > 25)
summary(model_postpub_int)
pretty_estimates(model_postpub_int)
```

--so we remove the interaction term and find the age effect to be significant unlike the sex effect. Not the effect direction: IGF1 increases with age in the young, and decreases with age in adult.

```{r}
ancova_postpub <- update(model_postpub_int, formula = . ~ sex + age)
summary(ancova_postpub)
pretty_estimates(ancova_postpub)
```

Again to visualise these parallel lines, we need to predict with the model:
```{r}
preds <- expand.grid(sex = c("Female", "Male"), age = 25:82, stringsAsFactors = FALSE) %>% 
	bind_cols(pred = predict(ancova_postpub, .))

ggplot(mapping = aes(x = age, colour = sex)) +
	geom_point(aes(y = log_sigf1), filter(juul, age > 25)) +
	geom_line(aes(y = pred), preds)
```

## Q5: Include BMI
```{r}
juul <- mutate(juul, bmi = weight / (height / 100)^2)
glm <- lm(log_sigf1 ~ age + sex + bmi, data = juul, subset = age > 25)
summary(glm) # note that 96 observation are deleted to missingness!
pretty_estimates(glm)
```

Always look at the missigness pattern! Note that I transpose it, as I find it to give a more intuitive output. 1’s mean observed value, 0’s mean missing value. Column titles indicate the number of patients with the pattern in that column. The bottom row shows the total number of variables with missing values in that column. The right-most column shows the number of patients with missing values for each variable. This is not the case for the plot which is in the original direction. 
```{r}
library(mice)
t(md.pattern(juul, rotate.names = TRUE))
```

